---
title: "remap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{remap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Remapping

There are two conventions of arguments among the underlying functions used by `rio`. Let's call them *Base Convention* and *"Tidy" Convention*.

| Convention | file location | selection of sheet | header      | examples                                                     |
|------------|---------------|--------------------|-------------|--------------------------------------------------------------|
| Base       | `file`        | `which`            | `header`    | `clipr::read_clip_tbl`                                       |
| "Tidy"     | `path`        | `sheet`            | `col_names` | `readxl::read_xlsx`, `readxl::read_xls`, `readODS::read_ods` |

`rio` can map Base Convention into "Tidy" Convention (but not vice versa).

```{r map1}
library(rio)
export(list("mtcars" = mtcars, "iris" = iris), "example.xlsx")
import("example.xlsx", which = "mtcars")
```

But you can still use the "Tidy" Convention, if the underlying function supports it.

```{r map2}
import("example.xlsx", sheet = "mtcars")
```

# Ellipsis or "dot dot dot"

Additional parameters are usually passed to the underlying function as ellipsis (`...`).

```{r map3}
## n_max is an argument of readxl::read_xlsx
import("example.xlsx", sheet = "iris", n_max = 10)
```

Parameters that the underlying function do not recognize are silently ignored by default.

```{r map4}
import("example.xlsx", sheet = "iris", n_max = 10, pizza = "pineapple")
```

If you don't like this behavior, please change the option `rio.ignoreunusedargs` to `FALSE`, i.e. `option(rio.ignoreunusedargs = FALSE)`.

```r
options(rio.ignoreunusedargs = FALSE)
import("example.xlsx", sheet = "iris", n_max = 10, pizza = "pineapple")
```

```{r map5, error = TRUE, echo = FALSE}
R.utils::withOptions({
    import("example.xlsx", sheet = "iris", n_max = 10, pizza = "pineapple")
}, rio.ignoreunusedargs = FALSE)
```

```{r, echo = FALSE, results = 'hide'}
unlink("example.xlsx")
```
