<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Import, Export, and Convert Data Files}
-->

# Import, Export, and Convert Data Files #

The idea behind **rio** is to simplify the process of importing data into R and exporting data from R. This process is, probably unnecessarily, extremely complex for beginning R users. Indeed, R supplies [an entire manual](http://cran.r-project.org/doc/manuals/r-release/R-data.html) describing the process of data import/export. And, despite all of that text, most of the packages described are (to varying degrees) out-of-date. Faster, simpler, packages with fewer dependencies have been created for many of the file types described in that document. **rio** aims to unify data I/O (importing and exporting) into two simple functions: `import` and `export` so that beginners (and experienced R users) never have to think twice (or even once) about the best way to read and write R data.

The core advantage of **rio** is that it makes assumptions that the user is probably willing to make. Specifically, **rio** uses the file extension of a file name to determine what kind of file it is. This is the same logic used by Windows OS, for example, in determining what application is associated with a given file type. By taking away the need to manually match a file type (which a beginner may not recognize) to a particular import or export function, **rio** allows almost all common data formats to be read with the same function.

By making import and export easy, it's an obvious next step to also use R as a simple data conversion utility. Transferring data files between various proprietary formats is always a pain and often expensive. The `convert` function therefore combines `import` and `export` to easily convert between file formats (thus providing a FOSS replacement for programs like [Stat/Transfer](https://www.stattransfer.com/) or [Sledgehammer](http://www.openmetadata.org/site/?page_id=1089)).

## Supported file formats ##

**rio** supports a variety of different file formats for import and export.

| Format | Import | Export |
| ------ | ------ | ------ |
| Tab-separated data (.tsv) | Yes | Yes |
| Comma-separated data (.csv) | Yes | Yes |
| CSVY (CSV + YAML metadata header) (.csvy) | Yes | No |
| Pipe-separated data (.psv) | Yes | Yes |
| Fixed-width format data (.fwf) | Yes | Yes |
| Serialized R objects (.rds) | Yes | Yes |
| Saved R objects (.RData) | Yes | Yes |
| JSON (.json) | Yes | Yes |
| Stata (.dta) | Yes | Yes |
| SPSS and SPSS portable | Yes (.sav and .por) | Yes (.sav) |
| "XBASE" database files (.dbf) | Yes | Yes |
| Excel (.xls) | Yes |  |
| Excel (.xlsx) | Yes | Yes |
| Weka Attribute-Relation File Format (.arff) | Yes | Yes |
| R syntax (.R) | Yes | Yes |
| Shallow XML documents (.xml) | Yes | Yes |
| SAS and SAS XPORT | Yes (.sas7bdat and .xpt) |  |
| Minitab (.mtp) | Yes |  |
| Epiinfo (.rec) | Yes |  |
| Systat (.syd) | Yes |  |
| Data Interchange Format (.dif) | Yes |  |
| OpenDocument Spreadsheet  (.ods) | Yes |  |
| Fortran data (no recognized extension) | Yes |  |
| Clipboard (default is tsv) | Yes (Mac and Windows) | Yes (Mac and Windows) |

Additionally, any format that is not supported by **rio** but that has a known R implementation will produce an informative error message pointing to a package and import or export function. Unrecognized formats will yield a simple "Unrecognized file format" error.

### Import ###

**rio** allows you to import files in almost any format using one, typically single-argument, function. `import` infers the file format from the file's extension and calls the appropriate data import function for you, returning a simple data.frame. This works for any for the formats listed above.

```{r, echo=FALSE, results='hide'}
library("rio")

export(iris, "iris.csv")
export(iris, "iris.rds")
export(iris, "iris.dta")
export(iris, "iris_noext", format = "csv")
```

```{r}
library("rio")

x <- import("iris.csv")
y <- import("iris.rds")
z <- import("iris.dta")

# confirm identical
all.equal(x, y, check.attributes = FALSE)
all.equal(x, z, check.attributes = FALSE)
```

If for some reason a file does not have an extension, or has a file extension that does not match its actual type, you can manually specify a file format to override the format inference step. For example, we can read in a CSV file that does not have a file extension by specifying `csv`:

```{r}
head(import("iris_noext", format = "csv"))
```


```{r, echo=FALSE, results='hide'}
unlink("iris.csv")
unlink("iris.rds")
unlink("iris.dta")
unlink("iris_noext")
```


### Export ###

The export capabilities of **rio** are somewhat more limited than the import capabilities, given the availability of different functions in various R packages and because import functions are often written to make use of data from other applications and it never seems to be a development priority to have functions to export to the formats used by other applications. That said, **rio** currently supports the following formats:


```{r}
library("rio")

export(iris, "iris.csv")
export(iris, "iris.rds")
export(iris, "iris.dta")
```

It is also easy to use `export` as part of an R pipeline (from magrittr or dplyr). For example, the following code uses `export` to save the results of a simple data transformation:

```{r}
library("magrittr")
mtcars %>% subset(hp > 100) %>%  aggregate(. ~ cyl + am, data = ., FUN = mean) %>% export(file = "mtcars2.dta")
```

```{r, echo=FALSE, results='hide'}
unlink("iris.csv")
unlink("iris.rds")
unlink("iris.dta")
unlink("mtcars2.dta")
```

## Convert ##

The `convert` function links `import` and `export` by constructing a dataframe from the imported file and immediately writing it back to disk. `convert` invisibly returns the file name of the exported file, so that it can be used to programmatically access the new file.

Because `convert` is just a thin wrapper for `import` and `export`, it is very easy to use. For example, we can convert 

```{r}
# create file to convert
export(iris, "iris.dta")

# convert Stata to SPSS
convert("iris.dta", "iris.sav")
```

`convert` also accepts lists of arguments for controlling import (`in_opts`) and export (`out_opts`). This can be useful for passing additional arguments to import or export methods. This could be useful, for example, for reading in a fixed-width format file and converting it to a comma-separated values file:

```{r}
# create an ambiguous file
fwf <- tempfile(fileext = ".fwf")
cat(file = fwf, "123456", "987654", sep = "\n")

# see two ways to read in the file
identical(import(fwf, widths = c(1,2,3)), import(fwf, widths = c(1,-2,3)))

# convert to CSV
convert(fwf, "fwf.csv", in_opts = list(widths = c(1,2,3)))
import("fwf.csv") # check conversion
```

```{r, echo=FALSE, results='hide'}
unlink("iris.dta")
unlink("iris.sav")
unlink("fwf.csv")
unlink(fwf)
```

It is also possible to use **rio** on the command-line by calling `Rscript` with the `-e` (expression) argument. For example, to convert a file from Stata (.dta) to comma-separated values (.csv), simply do the following:

```
Rscript -e "rio::convert('iris.dta', 'iris.csv')"
```
